How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2022-09-25 2.07-storytelling.Rmd
Add files via upload
# updating chapters

But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that is discussed above. Instead, it follows the formula that is set out below: **Introduction, Materials and Methods, Results and Discussion**. This style does not treat the experiment as a complication on the way to the story’s goal. The structure introduces the rationale behind conducting the experiment, then objectively explains the findings, and lastly discusses their meaning in relation to what is already known. From this study, I can spin the story (above), because I know what the endpoint is. But this storyis just one interpretation of the correlation that was found. [Correlation is not causation](#argument), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed. The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).

There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory. This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example, at conferences. The [presentations](#proposal-presentation) that tell a story and entertain are those we tend to remember. Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity. In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly. For example, don’t be boring, don’t [bamboozle](#style), write an active narrative (not in the passive tense), use the magic of threes, and many more.

@@

>>>>

But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that is discussed above. Instead, it follows the formula that is set out below: **Introduction, Materials and Methods, Results and Discussion**. This style does not treat the experiment as a complication on the way to the story’s goal. The structure introduces the rationale behind conducting the experiment, then objectively explains the findings, and lastly discusses their meaning in relation to what is already known. From this study, I can spin the story (above), because I know what the endpoint is. But this story is just one interpretation of the correlation that was found. [Correlation is not causation](#argument), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed. The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).

There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory. This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example, at conferences. The [presentations](#proposal-presentation) that tell a story and entertain are those we tend to remember. Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity. In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly. For example, don’t be boring, don’t [bamboozle](#style), write an active narrative (not in the [passive voice](#passive)), use the magic of threes, and many more.

@@

## Of babies, werewolves and silver bullets
Another take on the need for a narrative to engage the reader when writing science is provided in a [blog post](http://ecoevoevoeco.blogspot.com/2014/10/how-to-writepresent-science-baby.html) by Andrew Hendry. These unlikely characters in a scientific document act as a metaphor to help writers remember that when communicating their results they require a subject area to pull the reader in (everyone is interested in a cute baby), then to introduce the central problematic that needs to be solved, or the villain to the peace of the cute baby (the werewolf). Lastly, the silver bullet is a metaphor for the approach taken that will resolve the threats of the werewolf to the cute baby. Once you've laid out your study with this narrative, you should be able to go on to show exactly how your silver bullet results in a dead werewolf, or at least a horribly wounded werewolf that is a lesser threat to the cute baby.

### 2022-09-25 2.08-citations.Rmd
Add files via upload
# updating chapters

@@

>Measey et al. [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.

@@

>Measey et al. [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals. They did this by constructing GISS scores for all individuals in all groups.

@@

What about page numbers?
Sometimes you'll see a citation with a colon and page number after. This really only needs to be used if you are quoting specific text on a particular page:

@@

@@

However, it would be wrong to give a statement on amphibian decline and cite Measey et al. [-@measey2016global]. They did not study amphibian decline. Instead, you should read the papers that they cite [e.g. @wake2008are; @collins2009extinction; @pimm2014biodiversity], and read around those to find studies on amphibian decline that are appropriate for your context. This underlines one important aspect of choosing citations where the statement that you make relates directly to the study carried out in the citation.

@@

The primary literature consists of studies or experiments that are done in order to test a hypothesis. Secondary literature includes reviews, syntheses or meta-analyses. Primacy (see below) is important, but this depends on the space you have and whether the review contains all the information you need to cite. Sometimes reviews (especially meta-analyses) provide more information. It’s preferable to use primary literature, but sometimes reviews (or meta-analyses) are actually more expedient to use, especially if they are not the focus of your study. You can even cite both when relevant.

>>>>

But there may be a lot of papers that have found the same result, and you might want to only cite one as an example.

>Invasive amphibians have some serious impacts on their recipient environments [e.g. @measey2016global].

@@

>Measey and others [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.

@@

>Measey and colleagues [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals. They did this by constructing GISS scores for all individuals in all groups.

@@

### What about page numbers?
Sometimes you'll see a citation with a colon and page number after. This really only needs to be used if you are quoting specific text on a particular page:

@@

## The problems with brackets {#brackets}
Citations use brackets, and brackets are also used for other parts in writing. The problems with brackets often come when trying to combine citations that have brackets with the other parts of writing that also need the brackets. This is compounded when trying to do all this in [citation software](#referencesoftware) when you don't have complete control over it.

Most [citation software](#referencesoftware) allows for flexibility of how you will use your citation. In the above examples we have seen the full citation, a version where just the year is included in the brackets, some text occurs before (prefix) or after (suffix) the citation within the brackets, or even a combination of these:

 - The classic: [@measey2016global]
 - Omitting the authors: Measey et al. [-@measey2016global]
 - Using a prefix: [e.g. @measey2016global]
 - Using a suffix: [@measey2016global: 976]
 - A combination of the above: Measey et al. [-@measey2016global: 976]

[Citation software](#referencesoftware) handles producing these different ways by omitting the author, adding a prefix, or adding a suffix, respectively. However, the default will be to use the classic version, and you will need to alter this to one of the others in the software as you are citing it (i.e. as you add it to the document).

It is very common that people simply don't know how to do these different ways of citing within their [citation software](#referencesoftware). The easiest is to look up "how to add suffix" or whatever it is that you want to do in the respective manual or in your search engine. The result of not being sufficiently skilled in your [citation software](#referencesoftware), is that you end up with confused text - and usually a surfeit of brackets. The examples below are all mistakes that I regularly see in theses and manuscripts:

>Measey et al. [@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.

>Invasive amphibians have some serious impacts on their recipient environments (e.g. Measey et al. [-@measey2016global]).

>Invasive amphibians have some serious impacts on their recipient environments (e.g. [@measey2016global]).

The above examples are obviously wrong (i.e. obvious to an examiner or reviewer), and immediately suggest that the writer hasn't invested enough time in mastering the [citation software](#referencesoftware), or isn't aware that what they have produced is incorrect. I'd encourage you to take the time to correct these kinds of errors in a timely manor, or at least before submitting your thesis or manuscript.

### Avoid back to back bracketing

It is better to combine information into one set of continuous brackets instead of having back to back sets of brackets. You can do this by using a prefix or a suffix as below:

>Some invasive amphibians have been investigated insufficiently compared to their impact [e.g. Asian common toads:   @measey2016global].

Instead of back to back bracketing as follows:

>Some invasive amphibians have been investigated insufficiently compared to their impact (e.g. Asian common toads)[@measey2016global].

If you must have brackets nested within each other (and I would argue that this is very unusual), then rather use different types of brackets, such as square brackets inside round brackets. But make sure that you do this consistently throughout your thesis or manuscript.

@@

However, it would be wrong to give a statement on amphibian decline and cite Measey et al. [-@measey2016global]. They did not study amphibian decline. Instead, you should read the papers that they cite [e.g. @wake2008are; @collins2009extinction; @pimm2014biodiversity], and read around those to find studies on amphibian decline that are appropriate for your context. This kind of mis-citation has been called an 'empty citation' [@todd2007citing], and underlines one important aspect of choosing citations where the statement that you make relates directly to the study carried out in the citation.

Todd et al. [-@todd2007citing] looked at mis-citation rates in ecology, concluding that only around three quarters of papers were correctly cited. They found that 11% of citations were ambiguous, and 7% had citations that did not support the statement. Around 5% of citations were classed as empty - citing works that cite works with relevance - also known as 'lazy author syndrome'.

@@

The primary literature consists of studies or experiments that are done in order to test a hypothesis. Secondary literature includes reviews, syntheses or meta-analyses. Primacy (see below) is important, but this depends on the space you have and whether the review contains all the information you need to cite. Sometimes reviews (especially meta-analyses) provide more information. It’s preferable to use primary literature, but sometimes reviews (or meta-analyses) are actually more expedient to use, especially if they are not the focus of your study. You can even cite both when relevant. Todd et al. [-@todd2007citing] suggested that citing review articles might occur when postgraduate students do not have access to sufficient literature because of [pay-walls](#paywall).

### 2022-09-25 2.10-referencesortware.Rmd
Add files via upload
# updating chapters

Because published papers play such a fundamental role in your [reading and comprehension](#critical) of a subject, and [citations](#citations) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense. A reference manager is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies. Unsurprisingly, they become something that you may well use a lot during the course of your PhD, and it is worth thinking carefully about which one you want to use.

@@

>>>>

Because published papers play such a fundamental role in your [reading and comprehension](#critical) of a subject, and [citations](#citations) are pivotal in chapter or paper writing, putting these elements together in a good reference manager (aka citation software) makes a lot of sense. A reference manager is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies. Unsurprisingly, they become something that you may well use a lot during the course of your PhD, and it is worth thinking carefully about which one you want to use.

@@

## It's worth putting in the time to learn relevant parts of the package you choose

These days, reference managers are powerful and you might not need all of the functions that they have to offer. However, it is worth learning immediately what you will use, and this might be more than you think at first. For example, [using brackets](#brackets) is not as straightforward as it seems as you might often have to use a prefix of suffix inside the brackets. If you don't sort this out from the start, you might get into a mess that you forget to fix later.

### 2022-09-25 3.01-Title.Rmd
Add files via upload
# updating chapters

@@

@@

    -   Allow yourself a longer subtitle if needed, but don't go over 20 words total (some journals may limit your total to less).
    -   Consider the (former) 120 character limit of a Tweet as an upper limit

@@

-   Don't feel obliged to include [taxonomic terms](#taxonomy) unless it
    is relevant or compulsory

@@

-   Avoid obscure or specialist words that won't be understood by your
    readership

@@

-   Don't have your title as a question. Rather provide the answer!

>>>>

In general, there are three types of title:

-   Declarative - indicating the results of the work.

-   Descriptive - stating the research topic.

-   Interrogative - Asking a question.

Each title type has its pros and cons, but it might be worth starting by knowing which title type you want to achieve.

@@

If you are interested in reading more about titles and how people use them, there is some interesting analysis out there to get you started [e.g. @soler2007writing, @jamali2011article, @cook2016scholars, @krausman2020writing]. Remember that title type and use does vary by discipline, so it would be worth being aware of what is generally used in the journals that you intend to target.

@@

    -   Allow yourself a longer subtitle if needed, but don't go over 20 words total (some journals may limit your total to less). 
        -   Titles is a colon are downloaded and cited less [@jamali2011article].
    -   Artiles with longer titles are downloaded less [@jamali2011article].
    -   Consider the (former) 120 character limit of a Tweet as an upper limit

@@

-   Don't feel obliged to include [taxonomic terms](#taxonomy) unless it
    is relevant or compulsory [see @krausman2020writing]

@@

-   Avoid obscure or specialist words that won't be understood by your
    readership [@krausman2020writing]

@@

-   Don't use an implicit words in the title: a study of..., the effect of..., factors affecting..., etc. [@krausman2020writing].

-   Don't imply a series of works (e.g. Part I) [@krausman2020writing].

-   Don't be too funny. Sagi and Yechiam [-@sagi2008amusing] found that highly amusing titles got less citations.

-   Don't capitalise every word, the title is a sentence.

    -   Some journals will have a house style that may change your title to capitals, all caps or some other format. However, the title is a sentence, and sentence case makes it easier to read.

### Should your title have a question?

[Betteridge's Law](https://web.archive.org/web/20090226202006/http://www.technovia.co.uk/2009/02/techcrunch-irresponsible-journalism.html) relates to newspaper headlines and the idea that any headline that is a question is most simply answered by a single word "No" even though the writer has written the question with the expectation that readers will be interested in their explanation of why the answer is "Yes". In academic circles, this same idea is known as Hinchcliffe’s Rule: “If the title of a scholarly article is a yes–no question, the answer is ‘no’” [@shieber2015this]. The concern in both studies is that simple questions are being used as 'click-bait' to encourage potential readers to discover the simple answer. Indeed, there is some evidence that using a question in the title results in more downloads but less citations [@jamali2011article].

Ball [-@ball2009scholarly] found that the use of question marks in titles has been steadily rising in the life-sciences over 40 years. His assumption is that, like a newspaper headline, using a question in the title is a way of the authors trying to market the contents of their article. Cook and Plourde [-@cook2016scholars] analysed nearly 8000 titles (although not in the Biological Sciences) for Betteridge’s Law and Hinchcliffe’s Rule, but found that academics in the disciplines they analysed did not conform to either.

In summary, avoid having a simple question as your title.

### 2022-09-25 index.Rmd
Add files via upload
# updating chapters

--- 
title: 'How to write a PhD in Biological Sciences: a guide for the uninitiated'
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
csl: peerj.csl
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: "You don’t have to be a genius to write a PhD. It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules. This book is a guide specifically for writing a PhD in the biological sciences. It will guide you from deciding  whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences. The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals. Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
github-repo: yihui/bookdown-crc
graphics: yes
---

@@

# install the packages needed by this book; you fill out c(), e.g. c('ggplot2', 'dplyr')
lapply(c('xfun'), function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})
```
# In Press {-}
This book is _in press_ with CRC Press.

@@

You can pre-order your own copy on the [CRC Press website](https://www.routledge.com/How-to-Write-a-PhD-in-Biological-Sciences-A-Guide-for-the-Uninitiated/Measey/p/book/9781032080208), on [Amazon](https://www.amazon.com/How-Write-PhD-Biological-Sciences/dp/1032080205), at [Barns & Noble](https://www.barnesandnoble.com/w/how-to-write-a-phd-in-biological-sciences-john-measey/1139436164), and at other good book sellers.

@@

Before I go on, there are a great many people that I need to thank. First and foremost are my students, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. I have also been a student, and have been inspired by colleagues around the world who are exemplary advisors. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community (especially #academicTwitter) for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. James Baxter-Gilbert, Ellie Brown, Jack Dougherty, Anthony Herrel, Allan Ellis, Andrea Melotto, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, James Vonesh, Carla Wagener all read or commented on different aspects of the book. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.

>>>>

--- 
title: 'How to write a PhD in Biological Sciences: a guide for the uninitiated'
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
csl: peerj.csl
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: "You don’t have to be a genius to write a PhD. It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules. This book is a guide specifically for writing a PhD in the biological sciences. It will guide you from deciding  whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences. The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals. Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
url: howtowriteaphd.org
github-repo: How-to-write-a-PhD-in-Biological-Sciences/docs
graphics: yes
cover-image: figures/Cover-400wide.png
favicon: figures/favicon.png
---

@@

# install the packages needed by this book; you fill out c(), e.g. c('ggplot2', 'dplyr')
lapply(c('xfun'), function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})
```

# Preface {-}

*Published by CRC Press, Bota Raton, USA.*

@@

**Cover illustration**: Saint George and the Dragon was adapted by the author from an original work by Paolo Uccello (~1470). While the student (St George), supported by their laboratory (White Horse), is attempting to bring about
the end of their PhD (Dragon) using this Fool’s Guide (lance), the advisor (Princess) is leading the PhD by a leash from the darkness into the community towards the Scientific Project (town).

You can order your own copy on the [CRC Press website](https://www.routledge.com/How-to-Write-a-PhD-in-Biological-Sciences-A-Guide-for-the-Uninitiated/Measey/p/book/9781032080208), on [Amazon](https://www.amazon.com/How-Write-PhD-Biological-Sciences/dp/1032080205), at [Barns & Noble](https://www.barnesandnoble.com/w/how-to-write-a-phd-in-biological-sciences-john-measey/1139436164), and at other good book sellers.

@@

Before I go on, there are a great many people that I need to thank. First and foremost are my students, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. I have also been a student, and have been inspired by colleagues around the world who are exemplary advisors. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community (especially #academicTwitter) for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. James Baxter-Gilbert, Ellie Brown, Jack Dougherty, Anthony Herrel, Andrea Hurst, Allan Ellis, Joe McArthur, Andrea Melotto, Max Mühlenhaupt, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, Ben Stevenson, James Vonesh, Carla Wagener all read or commented on different aspects of the book. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.
